---
title: "SARFRAZ_HW1_410"
author: "Hussain Sarfraz"
date: "3/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(leaflet)
library(tidyverse)
library(ggmap)
library(rgdal)
```

# Homework 1 - 410

## Step 1: Setting working directory and uploading/editing excel file

* I used the `setwd()` function to set my working directory
* I then used the `read_csv()` function to read in the excel file that contained the information for each zip code (the name of the excel file is `US_map_data`)
* I then categorized the continuous variables in the `Risk_category` to **categorical** variables. I did this so I can fill in the map shape file colors with 4 single colors instead of a mix of 2 colors

```{r}
############# setting the working directory and loading state info. dataset
setwd("C:/Users/hussainsarfraz/Desktop/DATA 410")
US_map_data <- read_csv('place_data.csv')

US_map_data$Risk_category <- US_map_data$Risk
US_map_data$Risk_category[US_map_data$Risk_category<=24]  <- 1
US_map_data$Risk_category[US_map_data$Risk_category>24 & US_map_data$Risk_category<=49] <- 2
US_map_data$Risk_category[US_map_data$Risk_category>49 & US_map_data$Risk_category<=74] <- 3
US_map_data$Risk_category[US_map_data$Risk_category>74 & US_map_data$Risk_category<=98] <- 4
```

## Step 2: Reading in the shapefiles shared 

* I used the `readOGR()` function to read in the shape files that were shared with me. I am doing this to **merge** the zip codes in the shape files with the zip codes in the excel file. This would help in displaying the map data.

```{r}
############# Reading in the shapefiles shared with me (from census website)
states <- readOGR("Zipcodes_Poly", layer = "Zipcodes_Poly", encoding = "UTF-8")
# names(states@data) #this shows all the column names in the 'states' object
# head(states@data) #this shows the data values stored under each column. Gives us a brief look of the dataset

############# Merging the excel file with shape codes file through zip codes column
states@data <- data.frame(states@data, US_map_data[match(states@data$CODE, US_map_data$CODE),])
```

## Step 3: Creating the pop-up 

* I used the `paste0()` function to create the information that was displayed on the pop-up that is on my map. I stored this information in the `state_popup` object and will use this object later with creating my map in **step: 4**

```{r}
############# Creating the map pop-up 
state_popup <- paste0("<strong>Risk Index: </strong>",
                      states@data$Risk,"%",
                      '<br>',
                      "<strong>Party: </strong>", 
                      states@data$Poverty,"%", 
                      '<br>',
                      "<strong>Education: </strong>", 
                      states@data$Education,"%",
                      '<br>',
                      "<strong>Unemployment: </strong>",
                      states@data$Unemployment,"%",
                      '<br>',
                      "<strong>Crime: </strong>",
                      states@data$Crime,"%",
                      '<br>',
                      "<strong>ACEs: </strong>",
                      states@data$ACEs,"%")
```

## Step 4: Creating a color palette and map

* I used the `colorFactor()` function to create my color palette and specified the column that needs to be used for the color assignment. 
* I then made my map using `leaflet()`

```{r}
############# Creating a color scheme
factpal <- colorFactor(c("#ffdc87", "#ff9639","#ef404c","#d30020"),
                       states@data$Risk_category)
# factpal(c(1,2,3,4)) #used to check if the colors were correctly assigned to the variables used the 'Risk_category' column

############# Creating a map
leaflet(states) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(stroke = TRUE, 
              smoothFactor = 0.2, 
              fillOpacity = .8,
              color = 'black',
              fillColor = ~factpal(Risk_category),
              weight = 1, 
              popup = state_popup) %>%
  addLegend("bottomright", 
            colors =c(c("#ffdc87", "#ff9639","#ef404c","#d30020")),
            labels= c("0-24", "25-49" , "50-74" , "75-98"),  
            title= "Risk: Lowest to Highest",
            opacity = .8) 
```
