---
title: "SARFRAZ_FINAL_310"
author: "Hussain Sarfraz"
date: "3/5/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load('C:/Users/hussainsarfraz/Desktop/DATA 310/ANES20Datafor310.Rdata')

#Packages
library(survey)
library(stargazer)
library(sandwich)
library(tidyverse)
library(textutils)
```

# defining independent and dependent variables

```{r cars}
### defining independent and dependent variables ###
##Biden Feeling Thermometer - DEPENDENT VARIABLE
anes$biden.ft <- anes$V201151

anes$biden.ft[anes$biden.ft %in% c(-9,-4,998)] <- NA
summary(anes$biden.ft)

##BLM Movement Feeling Thermometer - INDEPENDENT VARIABLE
anes$blm.ft <- anes$V202174 

anes$blm.ft[anes$blm.ft %in% c(-9,-7,-6,-5,-4,998,999)] <- NA
summary(anes$blm.ft)
### creating new variables for difference of means 
#blm.ft
anes$blm.ft.category <- NA
anes$blm.ft.category[anes$blm.ft>=50] <- 1
anes$blm.ft.category[anes$blm.ft<50] <- 0

#blm.support
anes$blm.support<- NA
anes$blm.support[anes$blm.ft>=50] <- 1
anes$blm.support[is.na(anes$blm.support)] <- 0
# unique(anes$blm.support)

#blm.notsupport
anes$blm.notsupport<- NA
anes$blm.notsupport[anes$blm.ft<50] <- 1
anes$blm.notsupport[is.na(anes$blm.notsupport)] <- 0
# unique(anes$blm.support)

##Trump Feeling Thermometer - INDEPENDENT VARIABLE
anes$trump.ft <- anes$V201152

anes$trump.ft[anes$trump.ft %in% c(-9,-7,-6,-5,-4,998,999)] <- NA
summary(anes$trump.ft)

##Respondent liberal/conservative placement - INDEPENDENT VARIABLE
anes$respondent.LC.placement <- anes$V201200 

anes$respondent.LC.placement[anes$respondent.LC.placement %in% c(-9,-8,99)] <- NA
summary(anes$respondent.LC.placement)
#table(anes$respondent.LC.placement)

anes %>%
  filter(respondent.LC.placement == 4) %>%
  nrow() #1818 rows are being removed, converted to NA

### creating new variables for difference of means 
anes$liberal<- NA
anes$liberal[anes$respondent.LC.placement %in% c(1,2,3)] <- 1
anes$liberal[!(anes$respondent.LC.placement %in% c(1,2,3))] <- 0

anes$moderate<- NA
anes$moderate[anes$respondent.LC.placement %in% c(4)] <- 1
anes$moderate[!(anes$respondent.LC.placement %in% c(4))] <- 0

anes$conservative<- NA
anes$conservative[anes$respondent.LC.placement %in% c(5,6,7)] <- 1
anes$conservative[!(anes$respondent.LC.placement %in% c(5,6,7))] <- 0

##Respondent race/ethnicity - INDEPENDENT VARIABLE
anes$respondent.race <- anes$V201549x

anes$respondent.race[anes$respondent.race %in% c(-9,-8)] <- NA
summary(anes$respondent.race)
#table(anes$respondent.race)
### creating new variables for difference of means 
#res.race.category

anes$res.race.category <- NA
anes$res.race.category[anes$respondent.race %in% c(1)] <- 1
anes$res.race.category[anes$respondent.race %in% c(2,3,4,5,6)] <- 0
#anes$res.race.category[!(anes$respondent.race %in% c(1))] <- 0

unique(anes$res.race.category)
table(anes$res.race.category)

#res.white
anes$res.white <- NA
anes$res.white[anes$respondent.race %in% c(1)] <- 1
anes$res.white[is.na(anes$res.white)] <- 0

#res.nonwhite
anes$res.nonwhite <- NA
anes$res.nonwhite[anes$respondent.race %in% c(2,3,4,5,6)] <- 1
anes$res.nonwhite[is.na(anes$res.nonwhite)] <- 0

##Opinion on best way to deal with urban unrest and rioting - INDEPENDENT VARIABLE
anes$urban.unrest.opinion <- anes$V201429

anes$urban.unrest.opinion[anes$urban.unrest.opinion %in% c(-9,-8,99)] <- NA
summary(anes$urban.unrest.opinion)
#table(anes$urban.unrest.opinion)

##Opinion on building wall border with Mexico - INDEPENDENT VARIABLE
anes$mexico.border.opinion <- anes$V201424

anes$mexico.border.opinion[anes$mexico.border.opinion %in% c(-9,-8)] <- NA
summary(anes$mexico.border.opinion)
#table(anes$mexico.border.opinion)

##Opinion on U.S gov. policy towards unauthorized immigrants - INDEPENDENT VARIABLE
anes$us.immigrant.opinion <- anes$V201417

anes$us.immigrant.opinion[anes$us.immigrant.opinion %in% c(-9,-8)] <- NA
summary(anes$us.immigrant.opinion)
#table(anes$us.immigrant.opinion)

##Respondent age - INDEPENDENT VARIABLE
anes$age <- anes$V201507x

anes$age[anes$age==-9] <- NA
summary(anes$age)
#table(anes$age)

#Weight
anes$weight<- anes$V200010a

#Restrict to variables we care about
anes <- anes[,63:81]

#Create weight object
anes.w <- svydesign(ids = ~1,
                    data=anes,
                    weights = anes$weight)
```

# Describing Main Variables

```{r mylatextable, results = "asis"}
stargazer(subset(anes,
                 select = c(biden.ft,blm.ft,respondent.LC.placement,respondent.race,trump.ft,urban.unrest.opinion,mexico.border.opinion,us.immigrant.opinion,age)),
                 type = 'latex',
           title = "Summary Statistic of Variables",
          covariate.labels = c('Joe Biden Feeling Thermometer',
                               'Black Lives Matter Feeling Thermometer',
                               'Liberal/Conservative Scale',
                               'Respondents Self-Identified Race',
                               'Trump Feeling Thermometer',
                               'Urban Unrest Opinion',
                               'Mexico Border Opinion',
                               'U.S Immigration Opinion',
                               'Respondents Age'))



stargazer(anes, type = 'latex')

# stargazer(subset(anes, 
#                  alabama.court$race == 'B',
#                  select = c(amountremain,amountdue,amountpaid)),
#                  type = 'latex')

```

## Graph of main variables (Biden Feeling Thermometer and BLM)
```{r}
anes %>% 
  select(biden.ft, blm.ft) %>%
  gather(key = 'which.ft', value = 'values') %>%
  ggplot() + 
  geom_density(aes(x=values,color=which.ft,fill=which.ft),alpha=0.3) +
  scale_fill_manual(values = c( "#1b98e0","red"),
                    name = "Color Legend",
                    labels = c("Biden Feeling Thermometer", "BLM Feeling Thermometer")) +
  scale_color_manual(values = c( "#1b98e0","red"),
                    name = "Color Legend",
                    labels = c("Biden Feeling Thermometer", "BLM Feeling Thermometer")) +
  xlab("Feeling Thermometer Score") +
  ylab('Density') +
  ggtitle("Joe Biden and Black Lives Matter (BLM) Feeling Thermometers") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```

#Trump Feeling thermometer and BLM
```{r}
anes %>% 
  select(trump.ft, blm.ft) %>%
  gather(key = 'which.ft', value = 'values') %>%
  ggplot() + 
  geom_density(aes(x=values,color=which.ft,fill=which.ft),alpha=0.3) +
  scale_fill_manual(values = c( "purple","red"),
                    name = "Color Legend",
                    labels = c("Trump Feeling Thermometer", "BLM Feeling Thermometer")) +
  scale_color_manual(values = c( "purple","red"),
                    name = "Color Legend",
                    labels = c("Trump Feeling Thermometer", "BLM Feeling Thermometer")) +
  xlab("Feeling Thermometer Score") +
  ylab('Density') +
  ggtitle("Trump and Black Lives Matter (BLM) Feeling Thermometers") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```

#Graph of liberal-conservative scale
```{r}
anes %>%
  ggplot()+
  geom_bar(aes(x=respondent.LC.placement)) +
  labs(title="Respondents Self-Placement of being Liberal-Conservative",  
       x="Liberal-Conservative Scale", 
       y = "Number of Respondents") +
scale_x_continuous(breaks=c(1,2,3,4,5,6,7)) + 
  xlab(paste("\n<<< More Liberal", 
             spaces(50), "More Conservative >>>")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```

#Respondents Self-Identified Race
```{r}
anes %>%
  ggplot()+
  geom_bar(aes(x=respondent.race)) +
  labs(title="Respondents Self-Identified Race",  
       x="Respondents Race", 
       y = "Number of Respondents") +
  scale_x_continuous(breaks=c(1,2,3,4,5,6),
                     labels = c('White','Black','Hispanic','Asian',' Native American','Mix')) + 
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```


#performing 3 t-tests
```{r}
#biden.ft,blm.ft.category
t3 <- svyttest(biden.ft ~ blm.ft.category, anes.w)
t3

#Among conservatives (biden.ft, res.LC.category&res.race.category, blm.ft.category) - CONSERVATIVE, WHITE
anes.w.r3LW <- svydesign(ids = ~1,
                      data=anes[anes$conservative==1 & anes$res.white==1,],
                      weights = anes$weight[anes$conservative==1 & anes$res.white==1])
t3CW <- svyttest(biden.ft ~ blm.ft.category, anes.w.r3LW)
t3CW

#Among conservatives (biden.ft, res.LC.category&res.race.category, blm.ft.category) - CONSERVATIVE, NON-WHITE
anes.w.l3LW <- svydesign(ids = ~1,
                      data=anes[anes$conservative==1 & anes$res.nonwhite==1,],
                      weights = anes$weight[anes$conservative==1 & anes$res.nonwhite==1])
t3CNW <- svyttest(biden.ft ~ blm.ft.category, anes.w.l3LW)
t3CNW

#Among liberals (biden.ft, res.LC.category&res.race.category, blm.ft.category) - LIBERAL, WHITE
anes.w.l..3LW <- svydesign(ids = ~1,
                      data=anes[anes$liberal==1 & anes$res.white==1,],
                      weights = anes$weight[anes$liberal==1 & anes$res.white==1])

tl3LW <- svyttest(biden.ft ~ blm.ft.category, anes.w.l..3LW)
tl3LW

#Among liberals (biden.ft, res.LC.category&res.race.category, blm.ft.category) - LIBERAL, NON-WHITE
anes.w.l..3LW <- svydesign(ids = ~1,
                      data=anes[anes$liberal==1 & anes$res.nonwhite==1,],
                      weights = anes$weight[anes$liberal==1 & anes$res.nonwhite==1])

tl3LNW <- svyttest(biden.ft ~ blm.ft.category, anes.w.l..3LW)
tl3LNW
```


#calculating means, extreme difference and probabilities of variables
```{r}
#mean rating for individuals who do not support the BLM movement
round(weighted.mean(anes$biden.ft[anes$blm.ft.category==0], weight=anes$weight[anes$blm.ft.category==0], na.rm=T),2)
#mean rating for individuals who do support the BLM movement
round(weighted.mean(anes$biden.ft[anes$blm.ft.category==1], weight=anes$weight[anes$blm.ft.category==1], na.rm=T),2)
#probability of seeing extreme difference (48.4) of...
round(t3$statistic,2) 
#...in our sample if the true difference if the population was 0 is...
format.pval(t3$p.value, eps=.01, nsmall=2)

#extreme difference for conservative-WHITE
round(t3CW$statistic,2)
#probability of getting extreme difference if the population was 0 is...
format.pval(t3CW$p.value, eps=.01, nsmall=2)

#extreme difference for conservative-NONWHITE
round(t3CNW$statistic,2)
#probability of getting extreme difference for if the population was 0 is...
format.pval(t3CNW$p.value, eps=.01, nsmall=2)

#extreme difference for liberal-WHITE
round(tl3LW$statistic,2)
#probability of getting extreme difference for if the population was 0 is...
format.pval(tl3LW$p.value, eps=.01, nsmall=2)

#extreme difference for liberal-NONWHITE
round(tl3LNW$statistic,2)
#probability of getting extreme difference if the population was 0 is...
format.pval(tl3LNW$p.value, eps=.01, nsmall=2)

```


#creating 6 graphs to show the difference in means

```{r}
#biden.ft -- blm.ft.category  CATEGORY: res.LC.category & res.race.category
#Visualize Main 
plot(density(anes$biden.ft[anes$blm.ft.category==0], na.rm=T),
     lwd=2, col="forestgreen", ylim=c(0,0.04), xlim=c(0,100),
     main="Pooled")
points(density(anes$biden.ft[anes$blm.ft.category==1],na.=T), 
       lwd=2, col="purple", type="l")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==0], na.rm=T), lty=2, col="forestgreen")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==1], na.rm=T), lty=2, col="purple")
legend("topright", c("Does not Support BLM", "Does Support BLM"), lty=c(1,1), lwd=c(2,2), 
       col=c("forestgreen", "purple"))

#Visualize conservative -WHITE
plot(density(anes$biden.ft[anes$blm.ft.category==0 & anes$conservative==1 & anes$res.white == 1], na.rm=T),
     lwd=2, col="forestgreen", ylim=c(0,0.04), xlim=c(0,100),
     main="White Conservatives")
points(density(anes$biden.ft[anes$blm.ft.category==1 & anes$conservative==1 & anes$res.white == 1],na.=T), 
       lwd=2, col="purple", type="l")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==0 & anes$conservative==1 & anes$res.white == 1], na.rm=T), lty=2, col="forestgreen")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==1 & anes$conservative==1 & anes$res.white == 1], na.rm=T), lty=2, col="purple")
legend("topright", c("Does not Support BLM", "Does Support BLM"), lty=c(1,1), lwd=c(2,2), 
       col=c("forestgreen", "purple"),)

#Visualize conservative -NON-WHITE
plot(density(anes$biden.ft[anes$blm.ft.category==0 & anes$conservative==1 & anes$res.nonwhite == 1], na.rm=T),
     lwd=2, col="forestgreen", ylim=c(0,0.04), xlim=c(0,100),
     main="Non-White Conservatives")
points(density(anes$biden.ft[anes$blm.ft.category==1 & anes$conservative==1 & anes$res.nonwhite == 1],na.=T), 
       lwd=2, col="purple", type="l")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==0 & anes$conservative==1 & anes$res.nonwhite == 1], na.rm=T), lty=2, col="forestgreen")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==1 & anes$conservative==1 & anes$res.nonwhite == 1], na.rm=T), lty=2, col="purple")
legend("topright", c("Does not Support BLM", "Does Support BLM"), lty=c(1,1), lwd=c(2,2), 
       col=c("forestgreen", "purple"),)

#Visualize liberal -WHITE
plot(density(anes$biden.ft[anes$blm.ft.category==0 & anes$liberal==1 & anes$res.white == 1], na.rm=T),
     lwd=2, col="forestgreen", ylim=c(0,0.04), xlim=c(0,100),
     main="White Liberals")
points(density(anes$biden.ft[anes$blm.ft.category==1 & anes$liberal==1 & anes$res.white == 1],na.=T), 
       lwd=2, col="purple", type="l")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==0 & anes$liberal==1 & anes$res.white == 1], na.rm=T), lty=2, col="forestgreen")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==1 & anes$liberal==1 & anes$res.white == 1], na.rm=T), lty=2, col="purple")
legend("topright", c("Does not Support BLM", "Does Support BLM"), lty=c(1,1), lwd=c(2,2), 
       col=c("forestgreen", "purple"), bg="white")

#Visualize liberal -NON-WHITE
plot(density(anes$biden.ft[anes$blm.ft.category==0 & anes$liberal==1 & anes$res.nonwhite == 1], na.rm=T),
     lwd=2, col="forestgreen", ylim=c(0,0.04), xlim=c(0,100),
     main="Non-White Liberals")
points(density(anes$biden.ft[anes$blm.ft.category==1 & anes$liberal==1 & anes$res.nonwhite == 1],na.=T), 
       lwd=2, col="purple", type="l")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==0 & anes$liberal==1 & anes$res.nonwhite == 1], na.rm=T), lty=2, col="forestgreen")
abline(v=mean(anes$biden.ft[anes$blm.ft.category==1 & anes$liberal==1 & anes$res.nonwhite == 1], na.rm=T), lty=2, col="purple")
legend("topright", c("Does not Support BLM", "Does Support BLM"), lty=c(1,1), lwd=c(2,2), 
       col=c("forestgreen", "purple"), bg="white")
```


#setting up the regression

```{r}
#biden.ft, blm.ft
m1 <- lm(biden.ft ~  blm.ft, data=anes, weights= anes$weight)
m1.vcov <- vcovCL(m1, cluster=anes$state)
m1.se <- sqrt(diag(m1.vcov))


#biden.ft, respondent.LC.placement
m2 <- lm(biden.ft ~ respondent.LC.placement, data=anes, weights= anes$weight)
m2.vcov <- vcovCL(m2, cluster=anes$state)
m2.se <- sqrt(diag(m2.vcov))


#biden.ft, res.race.category
m3 <- lm(biden.ft ~ res.race.category, data=anes, weights= anes$weight)
m3.vcov <- vcovCL(m3, cluster=anes$state)
m3.se <- sqrt(diag(m3.vcov))

#biden.ft, blm.ft, respondent.LC.placement, res.race.category
m4 <- lm(biden.ft ~ blm.ft + respondent.LC.placement + res.race.category, data=anes, weights= anes$weight)
m4.vcov <- vcovCL(m3, cluster=anes$state)
m4.se <- sqrt(diag(m3.vcov))
```

#making the table
```{r, echo=F, results = "asis", out.width="100%"}
stargazer(m1, m2,m3, m4,
          se=list(m1.se,m2.se,m3.se,m4.se),
          dep.var.labels = "Biden Feeling Thermometer",
          covariate.labels = c("BLM Feeling Thermometer","Ideology", "Respondent Race"),
          title = "Direct Effects",
          digits=2,
          float=T,
          single.row = T,
          column.sep.width = "1pt",
          font.size = "small",
          header = F,
          style="ajps",
          omit.stat = c("rsq","f", "ser"),
          star.cutoffs = c(.05),
          notes.append = FALSE,
          notes = c("OLS regression weighted to US adult population..", "Robust standard errors clustered by state. *p<0.05"))
```

#performing regression 

```{r}
#biden.ft, blm.ft
summary(lm(biden.ft ~ blm.ft,
           data = anes,
           weight = anes$weight))

#biden.ft, respondent.LC.placement
summary(lm(biden.ft ~ respondent.LC.placement,
           data = anes,
           weight = anes$weight))

#biden.ft, respondent.race
summary(lm(biden.ft ~ respondent.race,
           data = anes,
           weight = anes$weight))

#biden.ft, blm.ft, respondent.LC.placement, respondent.race
summary(lm(biden.ft ~ blm.ft + respondent.LC.placement + respondent.race,
           data = anes,
           weight = anes$weight))



summary(lm(biden.ft ~ blm.ft + respondent.LC.placement + respondent.race + urban.unrest.opinion + mexico.border.opinion + us.immigrant.opinion + age,
           data = anes,
           weight = anes$weight))
```

#regression with interaction term

```{r}
m4 <- lm(biden.ft ~ blm.ft*res.race.category*respondent.LC.placement, data=anes, weights= anes$weight)
m5 <- lm(biden.ft ~ blm.ft*res.race.category*respondent.LC.placement + urban.unrest.opinion + mexico.border.opinion + us.immigrant.opinion + age, data=anes, weights= anes$weight)
```

#interaction term regression graph
```{r, echo=F, results = "asis", out.width="100%"}
stargazer(m4,m5,
          dep.var.labels = "Biden Feeling Thermometer",
          covariate.labels = c("BLM Feeling Thermometer","Race","Ideology","Urban Unrest Opinion",
                               "Mexico Border Opinion","US Immigration Opinion","Age","BLM*Race","BLM*Ideology","Race*Ideology","BLM*Race*Ideology"),
          title = "Interactive Effects",
          digits=2,
          float=T,
          single.row = T,
          column.sep.width = "1pt",
          font.size = "small",
          header = F,
          style="ajps",
          omit.stat = c("rsq","f", "ser"),
          star.cutoffs = c(.05),
          notes.append = FALSE,
          notes = c("OLS regression weighted to US adult population.", "Robust standard errors clustered by state. *p<0.05"))

#The probability of seeing an interaction effect this extreme if the true interaction effect in the population was 0 is < 0.01 
format.pval(summary(m4)$coefficients["blm.ft:res.race.category:respondent.LC.placement",4], eps=.01, nsmall=2)
```

#assessing the coefficiants

```{r}
ideo.levels <- seq(1,7)
#blm.ft:respondent.LC.placement
effect.blm <- coef(m4)["blm.ft"] + coef(m4)["blm.ft:respondent.LC.placement"]*ideo.levels
effect.blm
#blm.ft:respondent.LC.placement
effect.race <- coef(m4)["res.race.category"] + coef(m4)["res.race.category:respondent.LC.placement"]*ideo.levels
effect.race

two.lev <- seq(1,2)
#blm.ft:res.race.category
effect.ra1 <- coef(m4)["blm.ft"] + coef(m4)["blm.ft:res.race.category"]*two.lev
effect.ra1
```
#biden feeling thermometer and ideology

\begin{centering}
\begin{table}[]
\caption{Marginal Effect of BLM Support, by Ideology}
\begin{tabular}{l|ccccccc}
\textbf{Ideology}              & Very Liberal & 2    & 3    & Moderate    & 5     & 6     & Very Conservative \\
\hline
\textbf{BLM Support} & 0.74224032         & 0.61689218 & 0.49154403 & 0.36619589 & 0.24084774 & 0.11549960 & -0.00984855            
\end{tabular}
\end{table}
\end{centering}

#race and ideology

\begin{centering}
\begin{table}[]
\caption{Marginal Effect of Race, by Ideology}
\begin{tabular}{l|ccccccc}
\textbf{Ideology}              & Very Liberal & 2    & 3    & Moderate    & 5     & 6     & Very Conservative \\
\hline
\textbf{Effect of Race} & 1.576609         & -1.805567 & -5.187743 & -8.569920 & -11.952096 & -15.334272 & -18.716449            
\end{tabular}
\end{table}
\end{centering}

#BLM and race

\begin{centering}
\begin{table}[]
\caption{Marginal Effect of BLM Support, by Race}
\begin{tabular}{l|ccccccc}
\textbf{Ideology}              & Non-White & White  \\
\hline
\textbf{Effect of BLM} & 0.5735768         &  0.2191901
\end{tabular}
\end{table}
\end{centering}
