---
title: "SARFRAZ_HW_6"
author: "Hussain Sarfraz"
date: "2/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gridExtra)
```

# Question 1

## Part A

* I used the `load()` function to load in the `acs` dataset.
* I then divided the `median.income` variable by $1,000 so each number can be expressed in thousands of dollars. I did this so it could be easier to see the unit changes when I graph the `median.income` variable.

```{r}
load('C:/Users/hussainsarfraz/Desktop/DATA 310/Week 6/ACSCountyData (1).Rdata')
acs$median.income <- acs$median.income / 1000
```

## Part B

When I see the **Relationship between Median Income and Child Poverty** scatter plot I notice that the points take a downward slope shape as median income increases. The points do not decrease in a linear manner in the sense that they gradually decrease as the median income increases. 

From looking at the scatter plot I can see that as the median income increases the percentage of child poverty decreases which means that when people earn higher income the percentage of children in poverty declines.

```{r}
ggplot(data=acs) +
  geom_point(aes(x = median.income, y = percent.child.poverty)) + 
  xlab("Median Income (Thousands)") +
  ylab("Percentage of Child Poverty") +
  scale_x_continuous(breaks=c(0,25,50,75,100,125,150)) + 
  ggtitle("Relationship between Median Income and Child Poverty") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```

## Part C

**NOTE**: I am using the symbols in the math equation $y=mx+b$ to represent the bi-variate linear regression

Regression Values | What it Means
-------------     | -------------
$b=53.584098$     | y-intercept
$m=-0.613294$     | slope
$x$               | input  (independent variable)
$y$               | output (dependent variable)

When I ran the bi-variate regression in R `summary(lm(percent.child.poverty~median.income, data=acs))` I got the above results. I have gave a detailed description below about what each number means:

* The **y-intercept** means when $x=0$ (meaning when no body is earning a income) then the child poverty rate would be at **53.584098** 
* The **slope** means that for every unit increase in `median.income` the child poverty rate would decrease by **-0.613294** percent. This makes sense because as people earn more income there is a lower chance of children being in poverty because parents can provide for their kids education and well-being.

When looking at the scatter plot you can see that the numbers that the bi-variate regression function produced are correct since the y-intercept ($b$) appears to be close to **53.584098** and the slope ($m$) decreases the output of the regression line ($y$). That is why the regression line is decreasing. 

```{r}
summary(lm(percent.child.poverty~median.income, data=acs))

ggplot(data=acs,aes(x = median.income, y = percent.child.poverty)) +
  geom_point() + 
  geom_smooth(method=lm, formula = y ~ poly(x, 1), se = FALSE) +
  xlab("Median Income (Thousands)") +
  ylab("Percentage of Child Poverty") +
  scale_x_continuous(breaks=c(0,25,50,75,100,125,150)) + 
  ggtitle("Relationship between Median Income and Child Poverty") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```

## Part D

When looking at the bi-variate linear regression graph in **part c** I realized that the regression does not satisfy **assumption 2** of the **Gauss-Markov Theory**. 

Assumption 2 of the Gauss-Markov Theory states that the true model that generates our data is:
$Y_{i} = \alpha + \beta_{1}*x_{i1} + \beta_{2}*x_{i2} + ... + \beta_{k}*x_{ik} + \epsilon_{i}$

When looking at the bi-variate regression line in **part c** you could see that at around a median income between **80-100 thousand dollars** the regression line goes off in a tangent and goes further away from the data points in the plot. According to the Gauss-Markov Theory this is inaccurate since the model (mentioned above) generates a true representation of the data.

To make my bi-variate regression line an accurate representation of the data I would need to add the square of `median.income` to the **linear** bi-variate regression line by changing the value of `poly(x,1)` to `poly(x,2)`. This would make the bi-variate regression into a curve and it would surround more data points as opposed to the **linear** bi-variate regression line. 

The new **polynomial/curved** bi-variate regression line improves the model fit since the line now surrounds most of the data points. But when we use data to inform us about a lines function form we **increase** the risk of over-fitting our model fit. This is what has occurred with the **polynomial/curved** bi-variate regression line. Here are a few limitations of the **polynomial/curved** bi-variate regression line:

* The y-intercept is not accurately represented through the line.
* Around a median income of **100-150 thousand dollars** the slope of the bi-variate regression line increases which in not accurate since the increase in slope means that when people earn higher income the child poverty rate increase. This **conflicts** with what was shown at the beginning of the **polynomial/curved** bi-variate regression line where there was a **negative** slope.

There are also changes with $R^2$ in the **linear** bi-variate regression line and **polynomial/curved** bi-variate regression line models. 

$R^2$ Value | **linear** or **polynomial/curved** bi-variate regression line model
------------- | -------------
$R^2=0.5549$  | **linear** bi-variate regression line model
$R^2= 0.7039$ | **polynomial/curved** bi-variate regression line model

The $R^2$ value increased in the **polynomial/curved** bi-variate regression line model. This increase happened because when I added the square of `median.income` to the **linear** bi-variate regression line I made the regression line come closer to most of the data points. This makes both variables more correlated. The increase in $R^2$ means that `median.income` explains a greater percentage ($70.39\%$) of the variation in the `percent.child.poverty` variable. This implies that $20.61\%$ of the variation in the `percent.child.poverty` variable is not explained by the `median.income` of people.

```{r}
summary(lm(percent.child.poverty~poly(median.income,2,raw=T), data=acs))

ggplot(data=acs,aes(x = median.income, y = percent.child.poverty)) +
  geom_point() + 
  geom_smooth(method=lm, formula = y ~ poly(x, 2), se = FALSE) +
  xlab("Median Income (Thousands)") +
  ylab("Percentage of Child Poverty") +
  scale_x_continuous(breaks=c(0,25,50,75,100,125,150)) + 
  ggtitle("Relationship between Median Income and Child Poverty") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))
```

## Part E

At around $\$30k$ thousand dollars an effect of an additional $\$1000$ in median income would **decrease** the percentage of child poverty. This is because the slope of the polynomial/curved bi-variate regression line is **negative**.

At around $\$100k$ thousand dollars an effect of an additional $\$1000$ in median income would **increase** the percentage of child poverty. This is because the slope of the polynomial/curved bi-variate regression line is **positive**.

The difference in slope does make theoretical sense because as a I explained in **part D**: 

* When we use data to inform us about a lines function form we **increase** the risk of over-fitting our model fit.

## Part F

Correlation Value | Variable Names
-------------     | -------------
$-0.4990148$      | median.income     and unemployment.rate
$-0.7448909$      | median.income     and percent.child.poverty
$0.6814938$       | unemployment.rate and percent.child.poverty

The least correlated variables are **median.income and unemployment.rate** and the most correlated variables are **median.income and percent.child.poverty**.

**unemployment.rate and percent.child.poverty** do have a strong correlation since the correlation score is $68.15\%$.

Now **unemployment.rate** does have a strong correlation with the dependent variable **percent.child.poverty** but is does not have a strong correlation with the independent variable **median.income**. This means that because **unemployment.rate** does not have a strong correlation with another independent variable the y-intercept ($b$) would decrease and the coefficient/slope of `median.income` would decrease. 

I wanted to see the conflicting correlations of the `median.income` and `unemployment.rate` variables so I compared the regression lines of both graphs side by side (the dependent variable for both graphs was `percent.child.poverty`). After seeing how the regression lines for both graphs were going in opposite directions I could understand why the y-intercept ($b$) and the coefficient/slope of `median.income` would decrease after adding `unemployment.rate` to the regression. 

```{r}
#median.income - unemployment.rate
cor(acs$median.income, acs$unemployment.rate, use="pairwise.complete") #-0.4990148

#median.income - percent.child.poverty
cor(acs$median.income, acs$percent.child.poverty, use="pairwise.complete") #-0.7448909

#unemployment.rate - percent.child.poverty
cor(acs$unemployment.rate, acs$percent.child.poverty, use="pairwise.complete") #0.6814938

median.income_graph <- ggplot(data=acs,aes(x = median.income, y = percent.child.poverty)) +
  geom_point() + 
  geom_smooth(method=lm, formula = y ~ poly(x, 1), se = FALSE) +
  xlab("Median Income (Thousands)") +
  ylab("Percentage of Child Poverty") +
  scale_x_continuous(breaks=c(0,25,50,75,100,125,150)) + 
  ggtitle("Relationship between Median Income and Child Poverty") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))

unemployment.rate_graph <- ggplot(data=acs,aes(x = unemployment.rate, y = percent.child.poverty)) +
  geom_point() + 
  geom_smooth(method=lm, formula = y ~ poly(x, 1), se = FALSE) +
  xlab("Unemployment Rate") +
  ylab("Percentage of Child Poverty") +
  scale_x_continuous(breaks=c(0,25,50,75,100,125,150)) + 
  ggtitle("Relationship between Unemployment Rate and Child Poverty") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        plot.title = element_text(size = 10))

grid.arrange(median.income_graph, unemployment.rate_graph)
```

## Part G

When I ran the regression for `percent.child.poverty` and `median.income` in **part C** the slope($m$) of `median.income` was $-0.613294$ which meant that a unit increase in `median.income` would cause a $-0.613294$ decrease in `percent.child.poverty`.

However, now that I am doing a **multivariate regression** with 2 independent variables `median.income` and `unemployment.rate` I noticed that the slope($m$) of `median.income` has decreased by $0.169458$ ($0.613294 - 0.443836 = 0.169458$). The slope($m$) that I have now is $-0.443836$. Although mathematically this would look like a **increase** in the context of this problem the change in slope is a **decrease** and a **negative thing**. This is because the now slope means that a unit increase in `median.income` reduces child poverty by $-0.443836$ percent. This percentage is **lower** than the rate that I got in the **bi-variate regression** which is a bad thing since my coefficient is telling me that a persons increase in income has a less effect in decreasing child poverty. 

The decrease in slope for the `median.income` variable could also mean that the `unemployment.rate` variable has a greater effect in changing the child poverty rate. We could see this in **part F** with the strong correlations that `median.income` and `unemployment.rate` had with `percent.child.poverty`.

```{r}
summary(lm(percent.child.poverty ~ median.income + unemployment.rate, data = acs))
```

## Part H

There is a slight decrease in the `median.income` and `unemployment.rate` coefficients if you compare them with the coefficients from **part G**. The difference is not too big and the explanation for the decrease is the new variables that are introduced in the regression analysis might have had a bigger impact in determining the percentage of child poverty which decreased the coefficient of the `median.income` and `unemployment.rate` variables (basically what happened in **part F** with the decrease in the coefficient of `median.income` when the `unemployment.rate` variable was introduced in the regression). The table below highlights the difference:

Coefficient Difference (Part G - Part H) | Variable Name
-------------                            | -------------
$-0.443836 - (-0.397512) = -0.046324$    | median.income
$1.369099 - 1.193781 = 0.175318$         | unemployment.rate

The question stated that living in the south could be associated with lower average incomes and more child poverty. The results of the multivariate regression proved this point since the `census.regionsouth` variable has the **highest coefficient** of $3.546065$ out of all the other variables. This means that with every unit/population increase in southwest region of America would cause a $354.6065\%$ increase in child poverty. 

The `census.regionwest` region had the lowest coefficient of $62.2232\%$ when compared with the other regions. This means that there is less poverty in the west when compared to the northern and southern sides of the states. 

```{r}
#type is character, convert to factor
class(acs$census.region)

acs$census.region <- factor(acs$census.region)
summary(lm(percent.child.poverty ~ median.income + unemployment.rate +
             census.region, 
           data = acs))
```

## Part I

When adding the variable that says if a county is urban or not I noticed that there were many changes in the coefficients (some increased while others decreased. I have highlighted these changes in the table below.) 

**NOTE**: I am making these comparisons from the results I got in **Part H**

Coefficient Difference (Part I - Part H)     | Variable Name
-------------                                | -------------
$-0.397512 - (-0.419528) = 0.022016$         | median.income
$1.193781 - 1.156250    = 0.037531$         | unemployment.rate
$1.193781 - 1.413684    =-0.219903$         | census.regionnortheast
$3.546065 - 3.371122    = 0.174943$         | census.regionsouth      
$0.622232 - 0.677346    =-0.055114$         | census.regionwest

The coefficient for `population.densityTRUE` was $4.302792$ which was the highest coefficient among all the variables. This means that the a urban or rural county does play a huge difference in determining if there is greater child poverty in one particular county or not. In this case urban areas increase child poverty by $43.02\%$ The variables `unemployment.rate` and `census.regionsouth` had a increase in coefficients as compared to from **part H**. This means that if the unemployment rate and county's located in the south have a higher increase in child poverty, especially if the county is in a urban area.   

```{r}
acs$population.density <- (acs$population.density >= 1000)

#convert class to factor
class(acs$population.density)

acs$population.density <- factor(acs$population.density)

summary(lm(percent.child.poverty ~ median.income + unemployment.rate +
             census.region + population.density, 
           data = acs))
```